---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

```{r getdata}
if(!file.exists("./data")){dir.create("./data")}
unzip("activity.zip", exdir = "./data")
(files <- list.files("./data", full.names=T))
dat <- read.csv(files[1], na.strings="NA")

head(dat)
tail(dat)
str(dat)

### change the date variable into class=date
dat$date <- as.Date(as.character(dat$date), "%Y-%m-%d")

## change steps variable into numeric
dat$steps <- as.numeric(dat$steps)
str(dat)

```


## What is mean total number of steps taken per day?

For this part of the assignment, you can ignore the missing values in the dataset.

1. Make a histogram of the total number of steps taken each day

2. Calculate and report the mean and median total number of steps taken per day

```{r}
library(dplyr)

stepsPerDay <- 
        dat %>%
        group_by(date) %>%
        summarize(totalSteps = sum(steps))

```

```{r q1a, fig.height=6, fig.width=8}
with(stepsPerDay, hist(totalSteps, breaks = "FD", main ="Total number of steps taken per day", xlab = "steps/day", col = "thistle4"))
```

```{r q1b, results='asis'}
totalStepsPerDaysummary <- 
        stepsPerDay %>%
        summarize(mean = mean(totalSteps, na.rm=T),
                  median = median(totalSteps, na.rm=T))
        
```

The **mean** total number of steps per day is `r round(totalStepsPerDaysummary$mean, digits=2)`

The **median** total number of steps per day is `r totalStepsPerDaysummary$median`

## What is the average daily activity pattern?

1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
library(dplyr)

stepsPerInterval <- 
        dat %>%
        group_by(interval) %>%
        summarize(avgNumSteps = mean(steps, na.rm = T))

```


```{r q2a, fig.width=8, fig.height=4}
with(stepsPerInterval, plot(interval, avgNumSteps, xlab="5 minute intervals", ylab="Average number of steps taken", main="Time series plot of the average number of steps taken\n at each of the 2355 5-min-intervals", type="l"))

```


```{r}
maxSteps <- which.max(stepsPerInterval$avgNumSteps)
maxInterval <- stepsPerInterval$interval[maxSteps]

```

The **`r maxInterval`th** interval contains, on average across all the days, the maximum number of steps.


## Imputing missing values

1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r q3a}
totalNAs <- sum(is.na(dat))
```

There are in total `r totalNAs` missing values in the dataset.

2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

3. Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r q3bc}
### use average number of steps for each 5-min interval to replace missing values in the 'steps' variable

tblDat <- tbl_df(dat) %>%
        group_by(interval) %>%
        mutate(imputed = mean(steps, na.rm=T))

tblDat$steps[is.na(tblDat$steps)] <- tblDat$imputed[is.na(tblDat$steps)]
head(tblDat)
```

4. Make a histogram of the total number of steps taken each day and calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r}
stepsPerDayNew <- 
        tblDat %>%
        group_by(date) %>%
        summarize(totalSteps = sum(steps, na.rm = T))
```

```{r q3dfig, fig.height=4, fig.width=12}
par(mfrow=c(1,2))
with(stepsPerDayNew, hist(totalSteps, main ="Total number of steps per day\n (missing values imputed from mean per interval)", xlab = "steps/day", col = "thistle4"))
with(stepsPerDay, hist(totalSteps, main ="Total number of steps per day", xlab = "steps/day", col = "thistle4"))
```

```{r q3d}
stepsPerDayNewSummary <- summarize(stepsPerDayNew, mean = mean(totalSteps), median = median(totalSteps))
```

The mean and median total number of steps for this new data is `r round(stepsPerDayNewSummary$mean, digits=2)` and `r round(stepsPerDayNewSummary$median, digits=2)`. Imputing missing values does not change mean and median by much. When we impute missing data the mean gets closer to median.


## Are there differences in activity patterns between weekdays and weekends?

1. Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.


```{r}
tblDat2 <- select(tblDat, steps:interval) 
test <- weekdays(tblDat2$date) %in% c("Saturday", "Sunday")
tblDat2$dayType <- as.factor(ifelse(test, "weekend", "weekday"))
head(tblDat2) 
str(tblDat2$dayType)

### create new data grouped by interval and by dayType

tblDat3 <- tblDat2 %>%
        group_by(interval, dayType) %>%
        summarize(mean=mean(steps))

```

```{r fig.height = 8, fig.width=8}
par(mfrow = c(2, 1), mar = c(4, 6, 2, 1.5), oma = c(1, 1, 1, 1))
weekdayDat <- subset(tblDat3, dayType=="weekday")
weekendDat <- subset(tblDat3, dayType=="weekend")
with(weekdayDat, plot(interval, mean, xlab="5 minute intervals", ylab="Average number of\n steps taken", main="Average number of steps per interval on weekdays", type="l"))
with(weekendDat, plot(interval, mean, xlab="5 minute intervals", ylab="Average number of\n steps taken", main="Average number of steps per interval on weekends", type="l"))
```

